Analytical Statistics
======================

###Edre MA, DrPH

###`r as.character(Sys.Date())`

#objective 3: To determine the factors contributing to hypertension<br>
#we want to know first what contributes to systolic hypertension

----------------------------------------<br>
#Comparing numerical values: parametric
----------------------------------------

#we use same modified "hstat" dataset as in descriptive statistics session<br)
#inspect first our dataset, ensure all numerical as numeric, categorical as factors<br>
#active dataset->variables->show variables in active dataset

```{r}
#####Show variables in active data set#####
```

### str
```{r}
str(hstat)
```

#reorder our exercise variables first


```{r}
#####Reorder Factor Levels#####
hstat$exercise <- factor(hstat$exercise, levels=c('Low','Moderate','High'), ordered=TRUE)
```

#if our IV is categorical with 2 categories<br>
#example, we want to know if being male has any relationship with sbp<br>
#stat analysis->continuous variable->two sample t test<br>
#we want to visualize the comparison too

### independent sample t test

```{r}
#####Two-sample t-test#####
group.names <- NULL
group.means <- NULL
group.sds <- NULL
group.p <- NULL
res <- NULL
(res <- t.test(sbp~factor(sex), alternative='two.sided', conf.level=0.95, var.equal=TRUE, data=hstat))
windows(width=7, height=7); par(lwd=1, las=1, family="sans", cex=1, mgp=c(3.0,1,0))
bar.means <- tapply(hstat$sbp, factor(hstat$sex), mean, na.rm=TRUE)
bar.sds <- tapply(hstat$sbp, factor(hstat$sex), sd, na.rm=TRUE)
bar.sds <- ifelse(is.na(bar.sds), 0, bar.sds)
barx <- barplot(bar.means, ylim=c(ifelse(min(bar.means, na.rm=TRUE)>0, 0, min(bar.means-bar.sds, na.rm=TRUE)*1.2), max(bar.means+bar.sds, na.rm=TRUE)*1.2), 
  xlab="sex", ylab="sbp", axis.lty=1)
error.bar(barx, bar.means, bar.sds)
group.names <- c(group.names, "sex=Female")
group.means <- c(group.means, bar.means[1])
group.sds <- c(group.sds, bar.sds[1])
group.p <- c(group.p, signif(res$p.value,digits=3))
group.names <- c(group.names, "sex=Male")
group.means <- c(group.means, bar.means[2])
group.sds <- c(group.sds, bar.sds[2])
group.p <- c(group.p, "")
summary.ttest <- NULL
summary.ttest <- data.frame(mean=group.means, sd=group.sds, p.value=group.p)
rownames(summary.ttest) <- group.names
colnames(summary.ttest) <- gettext(domain="R-RcmdrPlugin.EZR",colnames(summary.ttest))
summary.ttest
```

#now we know sex has no effect on sbp in our study<br>
#we want to know now, does exercise have an effect (low,mod,high intensity)<br>
#one way ANOVA (assume equal variance, if not use welch ANOVA)<br>
#stat analysis->continuous variable->one-way ANOVA

```{r}
library(mvtnorm, pos=23)
```


```{r}
library(survival, pos=23)
```


```{r}
library(MASS, pos=23)
```


```{r}
library(TH.data, pos=23)
```


```{r}
library(multcomp, pos=23)
```



### One-Way Analysis of Variance: sbp ~ factor(exercise)

### pairwise.t.test
```{r}
#####One-way ANOVA#####
group.names <- NULL
group.means <- NULL
group.sds <- NULL
group.p <- NULL
AnovaModel.1 <- aov(sbp ~ factor(exercise), data=hstat, na.action=na.omit)
windows(width=7, height=7); par(lwd=1, las=1, family="sans", cex=1, mgp=c(3.0,1,0))
bar.means <- tapply(hstat$sbp, factor(hstat$exercise), mean, na.rm=TRUE)
bar.sds <- tapply(hstat$sbp, factor(hstat$exercise), sd, na.rm=TRUE)
error.bar <- function(x, y, upper, lower=upper, length=0.1,...){
if(length(x) != length(y) | length(y) !=length(lower) | length(lower) != length(upper))
	stop("vectors must be same length")
	arrows(x,y+upper, x, y-lower, angle=90, code=3, length=length, ...)
}
barx <- barplot(bar.means, ylim=c(ifelse(min(bar.means, na.rm=TRUE)>0, 0, min(bar.means-bar.sds, na.rm=TRUE)*1.2), max(bar.means+bar.sds, na.rm=TRUE)*1.2), 
  xlab="exercise", ylab="sbp", axis.lty=1)
error.bar(barx, bar.means, bar.sds)
group.names <- c(group.names, "exercise=Low")
group.means <- c(group.means, bar.means[1])
group.sds <- c(group.sds, bar.sds[1])
res <- NULL
res <- summary(lm(sbp ~ factor(exercise), data=hstat))
group.p <- c(group.p, signif(pf(res$fstatistic[1], res$fstatistic[2], res$fstatistic[3], lower.tail=FALSE), digits=3))
group.names <- c(group.names, "exercise=Moderate")
group.means <- c(group.means, bar.means[2])
group.sds <- c(group.sds, bar.sds[2])
group.p <- c(group.p, "")
group.names <- c(group.names, "exercise=High")
group.means <- c(group.means, bar.means[3])
group.sds <- c(group.sds, bar.sds[3])
group.p <- c(group.p, "")
summary(AnovaModel.1)
summary.anova <- NULL
summary.anova <- data.frame(mean=group.means, sd=group.sds, p.value=group.p)
rownames(summary.anova) <- group.names
colnames(summary.anova) <- gettext(domain="R-RcmdrPlugin.EZR",colnames(summary.anova))
summary.anova
pairwise.t.test(hstat$sbp, hstat$exercise, var.equal=TRUE, p.adj="bonferroni")
```

#Exercise adds benefit in sbp reduction<br>
#does sbp correlate with weight?
#stat analysis->continuous variable->test for Pearsons correlation

### pearson correlation with Scatterplot: sbp~wt

```{r}
#####Test for Pearson's correlation#####
windows(width=7, height=7); par(lwd=1, las=1, family="sans", cex=1, mgp=c(3.0,1,0))
scatterplot(sbp~wt, regLine=list(method=lm, lty=1), smooth=FALSE, boxplots='xy', data=hstat)
res <- NULL
(res <- cor.test(hstat$sbp, hstat$wt, alternative="two.sided", method="pearson"))
cat(gettext(domain="R-RcmdrPlugin.EZR", "correlation coefficient"), " = ", signif(res$estimate, digits=3), ", ", gettext(domain="R-RcmdrPlugin.EZR", "95% CI"), 
  " ", signif(res$conf.int[1],digits=3), "-", signif(res$conf.int[2],digits=3), ", ", gettext(domain="R-RcmdrPlugin.EZR", "p.value"), " = ", signif(res$p.value, 
  digits=3), "
", sep="")
```

#yes, the heavier the person, the higher the sbp<br>
#now you conducted a high-intensity interval training (HIIT) intervention<br>
#you want to measure pre and post HIIT effect on weight<br>
#stat analysis->continuous->paired t test

### paired t test 

```{r}
#####Paired t-test#####
(res <- t.test(hstat$wt, hstat$wt2, alternative='two.sided', conf.level=0.95, paired=TRUE))
mean1 <- mean(hstat$wt, na.rm=TRUE)
mean2 <- mean(hstat$wt2, na.rm=TRUE)
sd1 <- sd(hstat$wt, na.rm=TRUE)
sd2 <- sd(hstat$wt2, na.rm=TRUE)
summary.ttest <- NULL
summary.ttest <- data.frame(mean=c(mean1, mean2), sd=c(sd1, sd2), p.value=c(signif(res$p.value, digit=3),""))
rownames(summary.ttest) <- c("wt", "wt2")
colnames(summary.ttest) <- gettext(domain="R-RcmdrPlugin.EZR",colnames(summary.ttest))
summary.ttest
```

#Now we know that sbp is affected by weight.<br>
#exercise gives additional benefit to weight reduction<br>
#you are now concerned with the exercise giving effect on cardiovascular health<br>
#homocysteine (hcy) relates to cardiovascular heath from literature<br>
#what are the factors contributing to hcy level?

------------------------------------------<br>
#Comparing numerical values: non-parametric
------------------------------------------

#try transforming data into normal distribution by ln<br>
#active dataset->variables->logarithmic transformation<br>
#then check again for normality

### log
```{r}
#####Logarithmic transformation#####
hstat$hcyln <- log(hstat$hcy, base=exp(1))
#New variable hcyln was made.
```

```{r}
#####Kolmogorov-smirnov test for normal distribution#####
windows(width=7, height=7); par(lwd=1, las=1, family="sans", cex=1, mgp=c(3.0,1,0))
hist2(hstat$hcyln, freq=F, main="", xlab="hcyln", ylab="", col="darkgray")
curve(dnorm(x, mean=mean(hstat$hcyln[!is.na(hstat$hcyln)]), sd=sd(hstat$hcyln[!is.na(hstat$hcyln)])), add=T)
skewness.kurtosis(hstat$hcyln)
ks.test(hstat$hcyln, "pnorm", mean=mean(hstat$hcyln[!is.na(hstat$hcyln)]), sd=sd(hstat$hcyln[!is.na(hstat$hcyln)]))
# Shapiro-Wilk test can be performed only when the sample size is less than 5000. (Sample size = 153)
shapiro.test(hstat$hcyln)
```

#still not normally distributed<br>
#need to do non-parametric test

#female has higher or lower hcy level compared to male?<br>

### mann whitney U test

#stat analysis->nonparametric->mann whitney U test<br>

```{r}
#####Mann-Whitney U test#####
group.names <- NULL
group.median <- NULL
group.min <- NULL
group.max <- NULL
group.1Q <- NULL
group.3Q <- NULL
group.p <- NULL
windows(width=7, height=7); par(lwd=1, las=1, family="sans", cex=1, mgp=c(3.0,1,0))
boxplot(hcy~ factor(sex), ylab="hcy", xlab="sex", data=hstat)
(res <- wilcox.test(hcy ~ factor(sex), alternative="two.sided", data=hstat))
group.names <- c(group.names, "sex=Female")
group.min <- c(group.min, with(hstat, min(hcy[sex=='Female'], na.rm=TRUE)))
group.1Q <- c(group.1Q, with(hstat, quantile(hcy[sex=='Female'], 0.25, na.rm=TRUE)))
group.median <- c(group.median, with(hstat, median(hcy[sex=='Female'], na.rm=TRUE)))
group.3Q <- c(group.3Q, with(hstat, quantile(hcy[sex=='Female'], 0.75, na.rm=TRUE)))
group.max <- c(group.max, with(hstat, max(hcy[sex=='Female'], na.rm=TRUE)))
group.p <- c(group.p, signif(res$p.value,digits=3))
group.names <- c(group.names, "sex=Male")
group.min <- c(group.min, with(hstat, min(hcy[sex=='Male'], na.rm=TRUE)))
group.1Q <- c(group.1Q, with(hstat, quantile(hcy[sex=='Male'], 0.25, na.rm=TRUE)))
group.median <- c(group.median, with(hstat, median(hcy[sex=='Male'], na.rm=TRUE)))
group.3Q <- c(group.3Q, with(hstat, quantile(hcy[sex=='Male'], 0.75, na.rm=TRUE)))
group.max <- c(group.max, with(hstat, max(hcy[sex=='Male'], na.rm=TRUE)))
group.p <- c(group.p, "")
mannwhitney.table <- NULL
mannwhitney.table <- data.frame(Minimum=group.min, Q1=group.1Q, Median=group.median, Q3=group.3Q, Maximum=group.max, p.value=group.p)
rownames(mannwhitney.table) <- group.names
colnames(mannwhitney.table)[c(2,4)] <- c("25%", "75%")
colnames(mannwhitney.table) <- gettext(domain="R-RcmdrPlugin.EZR",colnames(mannwhitney.table))
mannwhitney.table
```

#sex has no signififant relationship with hcy<br>
#how about exercise intensity?

### kruskal wallis test

```{r}
#####Kruskal-Wallis test#####
windows(width=7, height=7); par(lwd=1, las=1, family="sans", cex=1, mgp=c(3.0,1,0))
boxplot(hcy~ factor(exercise), ylab="hcy", xlab="exercise", data=hstat)
tapply(hstat$hcy, hstat$exercise, median, na.rm=TRUE)
res <- NULL
(res <- kruskal.test(hcy ~ factor(exercise), data=hstat))
cat(gettext(domain="R-RcmdrPlugin.EZR", "Kruskal-Wallis test"), " ", gettext(domain="R-RcmdrPlugin.EZR", "p.value"), " = ", signif(res$p.value, digits=3), "
", sep="")
pairwise.kruskal.test(hstat$hcy, hstat$exercise, data.name="hstat", p.adjust.method="bonferroni")
```

#high intensity exercise significantly gives lower HCY compared to low/mod<br>
#you proceed in continuing the HIIT intervention as it gives benefit to both sbp and hcy<br>
#measure effectiveness again on weight reduction (wt2-wt3, assume non-normal wt3 data)

### wilcoxon signed rank test

```{r}
#####Wilcoxon's signed rank test#####
median(hstat$wt2 - hstat$wt3, na.rm=TRUE) # median difference
res <- NULL
(res <- wilcox.test(hstat$wt2, hstat$wt3, alternative='two.sided', paired=TRUE))
cat(gettext(domain="R-RcmdrPlugin.EZR", "Wilcoxon's signed rank test"), "p.value = 7.38e-27
")
```

#yes, the HIIT is effective in weight reduction<br>
#however you noticed some of your respondents are diabetic, risk of CVD is there<br>
#worry that your intervention gives more harm than good<br>
#finding relationship between hba1c and hcy<br>

### spearman correlation coefficient test

```{r}
#####Spearman's rank correlation test#####
windows(width=7, height=7); par(lwd=1, las=1, family="sans", cex=1, mgp=c(3.0,1,0))
scatterplot(hba1c~hcy, regLine=list(method=lm, lty=1), smooth=FALSE, boxplots='xy', data=hstat)
res <- NULL
(res <- cor.test(hstat$hba1c, hstat$hcy, alternative="two.sided", method="spearman"))
cat(gettext(domain="R-RcmdrPlugin.EZR", "Spearman's rank correlation coefficient"), signif(res$estimate, digits=3), gettext(domain="R-RcmdrPlugin.EZR", 
  "p.value"), " = ", signif(res$p.value, digits=3), "
")
```

#you conclude that hcy has no significant correlation with hba1c<br>
#in future, you would prioritize giving HITT intervention to the diabetic & hypertensive patients

#now, you are focused back to your objective 3<br>
#factors contributing to hypertension (hpt)

----------------------------<br>
#Comparing categorical variables
----------------------------

#assuming you are using same dataset with the transformed varibles(refer descriptive stat notes)<br>
#smoking has a relationship with hpt?<br>
#stat analysis -> discrete -> create two way table (tick "print expected frequencies" and "chi square test",untick "fisher exact")

### Chi square test

```{r}
#####Create two-way table and compare two proportions (Fisher's exact test)#####
Fisher.summary.table <- NULL
.Table <- NULL
.Table <- xtabs(~smoking+hpt, data=hstat)
.Table
rowPercents(.Table) # Row Percentages
.Test <- chisq.test(.Table, correct=TRUE)
.Test
.Test$expected # Expected Counts
remove(.Test)
res <- NULL
res <- chisq.test(.Table, correct=TRUE)
Fisher.summary.table <- rbind(Fisher.summary.table, summary.table.twoway(table=.Table, res=res))
colnames(Fisher.summary.table)[length(Fisher.summary.table)] <-  gettext(domain="R-RcmdrPlugin.EZR", 
  colnames(Fisher.summary.table)[length(Fisher.summary.table)])
Fisher.summary.table
```

#yes, smoking is significantly related to hpt. More smokers are hypertensive

#how about BMI status and hpt?

### Fisher's exact test

#do the same method as chi square, except now tick "fisher exact" box<br>
#used when more than 20% cells with expected count less than 5

```{r}
#####Create two-way table and compare two proportions (Fisher's exact test)#####
Fisher.summary.table <- NULL
.Table <- NULL
.Table <- xtabs(~bmistatus+hpt, data=hstat)
.Table
rowPercents(.Table) # Row Percentages
.Test <- chisq.test(.Table, correct=TRUE)
.Test
.Test$expected # Expected Counts
remove(.Test)
fisher.test(.Table)
res <- NULL
res <- fisher.test(.Table)
Fisher.summary.table <- rbind(Fisher.summary.table, summary.table.twoway(table=.Table, res=res))
colnames(Fisher.summary.table)[length(Fisher.summary.table)] <-  gettext(domain="R-RcmdrPlugin.EZR", 
  colnames(Fisher.summary.table)[length(Fisher.summary.table)])
Fisher.summary.table
```

#significant relationship between hpt and bmi status

---------------------------------------<br>
#Reporting your findings in table form
---------------------------------------

#example is the fisher exact test result for bmistatus-hpt just now<br>
#graphs and tables -> summary table of results -> two-way table+CSV as output

### table 1: association between BMI status and HPT

```{r}
#####Summary table of results#####
if(objectCheck("Fisher.summary.table", objects())) w.twoway(Fisher.summary.table, 
  filename="C:/Users/raef/Desktop/rcommander wshop/old files/IIUM_Biostatistic_workshop-main/Fisher.summary.table.csv", en=1)
```

#THANK YOU
