Descriptive Statistics
======================

###Edre MA, DrPH

###`r as.character(Sys.Date())`

#we will be using R commander to ease the learning<br>
#now lets start by looking at a scenario<br>
#You are a researcher involved in a hypertension study<br>
#objective 1: To describe the background characteristics of respondents<br>
#objective 2: To determine the prevalence of hypertension<br>
#objective 3: To determine the factors contributing to hypertension

#----------------------------------<br>
#Packages,library & setup<br>
#----------------------------------

#packages include "Rcdmr" AND "RcmdrPlugin.EZR"<br>
#first, install both packages <br>
install.packages("Rcmdr", repos='http://cran.us.r-project.org')<br>
install.packages("RcmdrPlugin.EZR", repos='http://cran.us.r-project.org')<br>
#load the package -> start the GUI version of R !<br>
library(Rcmdr)<br>
#go ahead, open R commander window and explore it<br>
#it looks pretty much like spss, suitable for beginners. and its free!<br>
#bit more setup<br>
#go to tool->option->output->adjust window height line to "10" AND untick "send output to R console"<br>
#restart R commander<br>
#then go back to tools -> load Rcmdr plugins -> load the "RcmdrPlugin.EZR" ->restart again r commander<br>
#save your settings (workspace). Save often

#--------------<br>
#data<br>
#--------------

#pulling the data from GitHub

#go to https://github.com/adilzainal/IIUM_Biostatistic_workshop<br>
#click "code" -> "Download ZIP"<br>
#extract the ZIP file using WinRAR<br>
#Create a new specific folder to store all files in your desktop<br>
#set as working directory in r commander<br>
#go to file -> change working directory

```{r}
setwd("C:/Users/raef/Desktop/rcommander wshop")
```

#loading the data

#if excel spreadsheet file (.xlsx)<br>
#file->import data-> from excel dataset ->rename as "hstat"->ok<br>
```{r}
hstat <- readXL("C:/Users/raef/Desktop/rcommander wshop/healthstatus6.xlsx", rownames=FALSE, header=TRUE, na="", sheet="healthstatus6", stringsAsFactors=TRUE)
```

#view data set<br>
#edit data set<br>

#----------------------------<br>
#objective 1: To describe the background characteristics of respondents<br>
#summarising numerical values<br>
#----------------------------

#we choose 3 IVs: age,sbp,dbp<br>
#statistical analysis->continuous variables->numerical summaries

### numSummary2
```{r}
#####Numerical summaries#####
res <- numSummary2(hstat[,c("age", "dbp", "sbp")], statistics=c("mean", "u.sd", "quantiles"), quantiles=c(0,.25,.5,.75,1))
colnames(res$table) <- gettext(domain="R-RcmdrPlugin.EZR", colnames(res$table))
res
```

#--------------------------<br>
#normality assumption check<br>
#--------------------------<br>
#there are 5 criteria before you make decision what to report:

#1.mean~median<br>
#select only "mean" and ".5" quantile

### numSummary2
```{r}
#####Numerical summaries#####
res <- numSummary2(hstat[,c("age", "dbp", "sbp")], statistics=c("mean", "quantiles"), quantiles=c(.5))
colnames(res$table) <- gettext(domain="R-RcmdrPlugin.EZR", colnames(res$table))
res
```

#2. qqplot<br>
#go to graph and tables->quantile-comparison plot

### Residual Quantile-Comparison Plot: hstat: age
```{r}
#####Quantile-comparison plot#####
windows(width=7, height=7); par(lwd=1, las=1, family="sans", cex=1, mgp=c(3.0,1,0))
qqPlot(hstat$age, dist="norm")
```

#3-5. bell shaped curve, skewness/kurtosis & normality test<br>
#statistical analysis->continuous variables->KS test for normal distribution<br>
#shapiro wilk for <5000, KS test for 5000 or more samples

### hist2

### curve

### skewness.kurtosis

### ks.test

### shapiro.test
```{r}
#####Kolmogorov-smirnov test for normal distribution#####
windows(width=7, height=7); par(lwd=1, las=1, family="sans", cex=1, mgp=c(3.0,1,0))
hist2(hstat$age, freq=F, main="", xlab="age", ylab="", col="darkgray")
curve(dnorm(x, mean=mean(hstat$age[!is.na(hstat$age)]), sd=sd(hstat$age[!is.na(hstat$age)])), add=T)
skewness.kurtosis(hstat$age)
ks.test(hstat$age, "pnorm", mean=mean(hstat$age[!is.na(hstat$age)]), sd=sd(hstat$age[!is.na(hstat$age)]))
# Shapiro-Wilk test can be performed only when the sample size is less than 5000. (Sample size = 153)
shapiro.test(hstat$age)
```

#finally, make your decision


### numSummary2
```{r}
#####Numerical summaries#####
res <- numSummary2(hstat[,c("age", "dbp", "sbp")], statistics=c("mean", "u.sd"), quantiles=c(0,.25,.5,.75,1))
colnames(res$table) <- gettext(domain="R-RcmdrPlugin.EZR", colnames(res$table))
res
```

#------------------------------<br>
#summarising categorical values<br>
#------------------------------

#go to statistical analysis->discrete variables->frequency distribution->tick show missing data

### Frequencies: exercise, sex, smoking

```{r}
#####Frequency Distributions#####
.Table <- NULL
(.Table <- table(hstat$exercise, exclude=NULL))  # counts for exercise
round(100*.Table/sum(.Table), 2)  # percentages for   exercise
.Table <- NULL
(.Table <- table(hstat$sex, exclude=NULL))  # counts for sex
round(100*.Table/sum(.Table), 2)  # percentages for   sex
.Table <- NULL
(.Table <- table(hstat$smoking, exclude=NULL))  # counts for smoking
round(100*.Table/sum(.Table), 2)  # percentages for   smoking
```

#count refers to the frequency of the category

#-------------<br>
#missing data<br>
#-------------

#usually coded as "NA" in the dataset<br>
#we create a dummy dataset "hstat2" first to showcase this exercise<br>
#file->import data->from excel file->rename to hstat2<br>
#click "edit data set" and delete "smoking" and "ht" entry for id5<br>\
#active data set-> missing observation-> count missing observation of all variables

### Count Missing Cases: hstat2
```{r}
sapply(hstat2, function(x)(sum(is.na(x)))) # NA counts
```

#-----------------<br>
#outlier detection<br>
#-----------------

#create an outlier dummy data <br>
#change id2 age to 499 for "hstat2" data and save
#check with statistical analysis ->continuous variables-> smirnov-grubbs test for outliers

### RemoveOutlier
```{r}
#####Smirnov-Grubbs test for outliers#####
RemoveOutlier(hstat2$age, return=0)
```

#visual method by boxplot


### bxp
```{r}
#####Boxplot#####
windows(width=7, height=7); par(lwd=1, las=1, family="sans", cex=1, mgp=c(3.0,1,0))
boxdata <- boxplot(hstat2$age[complete.cases(hstat2$age)], ylab="age", plot=FALSE)
boxdata$stats[1,] <- quantile(hstat2$age[complete.cases(hstat2$age)], .1, na.rm=TRUE)
boxdata$stats[5,] <- quantile(hstat2$age[complete.cases(hstat2$age)], .9, na.rm=TRUE)
boxdata.outliers <- hstat2$age[complete.cases(hstat2$age)][hstat2$age[complete.cases(hstat2$age)]<boxdata$stats[1,] | 
  hstat2$age[complete.cases(hstat2$age)]>boxdata$stats[5,]]
boxdata$out <- c(boxdata$out, boxdata.outliers)
boxdata$group <- c(boxdata$group, rep(1, length(boxdata.outliers)))
bxp(boxdata, ylab="age")
remove(boxdata)
remove(boxdata.outliers)
```


#--------------------------------------<br>
#objective 2: To determine the prevalence of hypertension<br>
#objective 3: To determine the factors contributing to hypertension<br>
#basic data transformation:categorizing<br>
#--------------------------------------


#hypertension status (either sbp or dbp equal or more than 140/90mmHg, respectively, considered hypertensive)<br>
#to answer objective 2, change back dataset to "hstat"<br>
#active dataset->variables->create new variable

```{r}
#####Create new variable#####
hstat$hpt <- with(hstat, ifelse(sbp<140 & dbp<90, 'normal','high'))
#New variable hpt was made.
```

#View or edit data set, we have the new variable "hpt" !<br>
#make it as a factor first (active dataset->variables->convert all character variables to factor)

### ChrToFactor
```{r}
#####Convert all character variables to factors#####
hstat <- ChrToFactor(hstat)
```

#prevalence : statistical analysis->discrete variables->frequency distributions

```{r}
#####Frequency Distributions#####
.Table <- NULL
(.Table <- table(hstat$hpt, exclude=NULL))  # counts for hpt
round(100*.Table/sum(.Table), 2)  # percentages for   hpt
```

#to make data preparation for objective 3<br>
#glucose control (6.5% and above considered poor)<br>
#active dataset->variables->recode variable

```{r}
#####Recode variables#####
hstat <- within(hstat, {
  glucontrol <- Recode(hba1c, '0.01:6.49="good"; 6.50:50.0="poor";', as.factor=TRUE)
})
```

#view your data to double check<br>

#bmistatus (WHO classification) : create new variable

```{r}
#####Create new variable#####
hstat$bmi <- with(hstat, wt / ((ht/100)^2))
#New variable bmi was made.
```

#double check again hstat before categorize<br>
#we use bin numeric variables with more than 2 groups

### cut
```{r}
#####Bin numeric variable to more than 2 groups with specified thresholds#####
hstat$bmistatus <- cut(hstat$bmi, breaks=c(-Inf, 18.50, 25.00, 30.00, Inf), labels=c("underweight", "normal", "overweight", "obese"), right=FALSE)
#New variable bmistatus was made.
table(hstat$bmistatus, exclude=NULL)
```

#recheck again bmistatus

#-------------------------------------------------<br>
#Reporting your descriptive analysis<br>
#-------------------------------------------------

#graphs and tables->summary table of sample characteristics

### finaltable_dataframe_print
```{r}
#####Summary table of sample characteristics#####
CatTable <- CreateCatTable(vars = c("bmistatus", "exercise", "glucontrol", "hpt", "sex", "smoking"), data=hstat)
matCatTable <- print(CatTable, printToggle = FALSE, showAllLevels = TRUE, exact=c("bmistatus", "exercise", "glucontrol", "hpt", "sex", "smoking"), smd = FALSE)
matCatTable <- matCatTable[,colnames(matCatTable)!="test"]
matCatTable <- cbind(Factor=row.names(matCatTable), matCatTable)
ContTable <- CreateContTable(vars = c("age", "dbp", "sbp"), data=hstat)
matContTable <- print(ContTable, printToggle = FALSE, explain = FALSE, smd = FALSE)
matContTable <- matContTable[,colnames(matContTable)!="test"]
matContTable <- cbind(level="", matContTable)
matContTable <- cbind(Factor=row.names(matContTable), matContTable)
ContnonnormalTable <- CreateContTable(vars = c("hcy"), data=hstat)
matContnonnormalTable <- print(ContnonnormalTable, printToggle = FALSE, nonnormal = TRUE, explain = FALSE, minMax=FALSE, smd = FALSE)
matContnonnormalTable <- matContnonnormalTable[,colnames(matContnonnormalTable)!="test"]
matContnonnormalTable <- cbind(level="", matContnonnormalTable)
matContnonnormalTable <- cbind(Factor=row.names(matContnonnormalTable), matContnonnormalTable)
FinalTable <- as.matrix(matCatTable)
tempStrata <- attributes(FinalTable)[[2]][2]
FinalTable <- rbind(FinalTable, matContTable)
FinalTable <- rbind(FinalTable, matContnonnormalTable)
attributes(FinalTable) <- c(list(dim=attributes(FinalTable)[[1]]), list(dimnames=c(attributes(FinalTable)[[2]][1], tempStrata)))
colnames(FinalTable)[2] <- "Group"
row0 <- colnames(FinalTable)
row1 <- FinalTable[1,]
row1 <- matrix(row1, nrow=1)
colnames(row1) <- row0
FinalTable <- FinalTable[which(rownames(FinalTable)!="n"),]
FinalTable <- rbind(n=row1, FinalTable)
FinalTable <- rbind(row0, FinalTable)
finaltable_dataframe_print(FinalTable)
write.table(FinalTable, file="C:/Users/raef/Desktop/rcommander wshop/tableone_descriptive.csv", sep=",", row.names=FALSE, col.names=FALSE)
```

#thank you



